---
title: 'Final Project for Biostat 202: Data Cleaning'
output: html_document
---

```{r setup, echo=FALSE}
```
# Introduction
## Background 
Acute Renal Failure (ARF) after surgery is a severe complication and it affects patients’ outcomes and hospital length. Especially, cardiovascular surgery is the second most likely to cause ARF following transplant surgery. Some biomarkers have been developed to predict ARF after surgery but have not been practical, and more easily access predict markers are desired in a clinical setting. American College of Surgeons National Surgical Quality Improvement Program® (ACS NSQIP®) data is developed by surgeons and includes patients’ outcomes that are thirty days after their operations. The purpose of this study is to predict ARF after cardiac surgery by pre-operation lab tests and conditions of operation using the NSQIP data.

## Research question 
Do pre-operation lab tests and conditions of operation predict Acute Renal Failure after surgery among people with cardiovascular surgery? 

## Variables/Predictors 
44 variables that are easily accessible by the time the operation is complete are included. We exclude weight, height, and days from lab tests to the operation because these variables are replaced by new variables, BMI, and the lab results within 14 days. 

Subjects' background: Sex, Age, BMI, Race and ethnicity, Smoking history, Comorbidities such as diabetes and congestive heart failure.
Lab results: BUN, serum creatinine, serum albumin, total bilirubin, and so on. The only lab results which are conducted within 14 days before the surgery are included.
Conditions of operation: Duration of Anesthesia, total operation time, and so on.

# Data preparation

*1. Import packages*

```{r}
library(tidyverse)
library(gt)
library(webshot2)
library(gtsummary)
```

*2. Load the dataset*

```{r data}
NSQIP <- read_csv(file = "Project 4 - NSQIP Data.csv")
```

*3. Exclude the subjects without cardiovascular surgery or with severe ARF with dialysis at the pre-operation time.*

```{r}
# Filter the subject with Cardiovascular Surgery by CPT code
dt_heart <- NSQIP %>%
  filter(33016 <= CPT & CPT <= 33999)
dt_arteries <- NSQIP%>%
  filter(34001 <= CPT & CPT<= 37799)
dt_cv <- rbind(dt_heart, dt_arteries)


# Exclude the patients with Dialysis at pre-op
dt_cv_no_dialysis <- dt_cv %>%
  filter(DIALYSIS != "Yes")


```

*4. The lab results which are conducted within 14 days before the surgery are included. This is because the lab results performed more than 15 days before surgery do not reflect the subject's condition at the preoperative timing. *

```{r code}
# Change the lab results to NA if the lab test was conducted before 15 days or more from the surgery. (DPRxxx: Days from the pre-lab test to the surgery) 
dt_lab_updated_bmi <- dt_cv_no_dialysis %>%
  mutate(
    PRSODM1 = if_else(DPRNA > 14 | is.na(DPRNA), NA, PRSODM),
    PRBUN1 = if_else(DPRBUN > 14 | is.na(DPRBUN), NA, PRBUN),
    PRCREAT1 = if_else(DPRCREAT > 14 | is.na(DPRCREAT), NA, PRCREAT),
    PRALBUM1 = if_else(DPRALBUM > 14 | is.na(DPRALBUM), NA, PRALBUM),
    PRBILI1 = if_else(DPRBILI > 14 | is.na(DPRBILI), NA, PRBILI),
    PRSGOT1 = if_else(DPRSGOT > 14 | is.na(DPRSGOT), NA, PRSGOT),
    PRALKPH1 = if_else(DPRALKPH > 14 | is.na(DPRALKPH), NA, PRALKPH),
    PRWBC1 = if_else(DPRWBC > 14 | is.na(DPRWBC), NA, PRWBC),
    PRHCT1 = if_else(DPRHCT > 14 | is.na(DPRHCT), NA, PRHCT),
    PRPLATE1 = if_else(DPRPLATE > 14 | is.na(DPRPLATE), NA, PRPLATE),
    PRPTT1 = if_else(DPRPTT > 14 | is.na(DPRPTT), NA, PRPTT),
    PRINR1 = if_else(DPRINR > 14 | is.na(DPRINR), NA, PRINR)
  )%>%
  # Calculate BMI
  mutate(
    BMI = WEIGHT/(HEIGHT^2)*703
  )

# Remove the duplicated cases
dt_unique_cases <- dt_lab_updated_bmi[duplicated(dt_lab_updated_bmi$CaseID)==FALSE,]
```


*5. Set the target*

```{r}
dt_target <- dt_unique_cases%>%
  mutate(renaloutcome = ifelse(RENAINSF=="Progressive Renal Insufficiency"|OPRENAFL=="Acute Renal Failure",1,0))
table(dt_target$renaloutcome)
write_csv(dt_target, "NSQIPdataset_cleaned.csv")

```

*6. Change the columns' names*

```{r}
dt_renamed <- dt_target %>%
  rename("Sex" = "SEX", "Race"="RACE_NEW",
         "Ethnicity"="ETHNICITY_HISPANIC" , "Smoke"="SMOKE",
         "Diabetes"="DIABETES" ,"COPD" ="HXCOPD",
         "Dyspnea"="DYSPNEA", "Congestive Heart Failure" ="HXCHF",
         "Hypertension"="HYPERMED", "Ascites"="ASCITES", 
         "Disseminated Cancer"="DISCANCR", "Bleeding disorders"="BLEEDDIS",
         "Severe Acure Renal Failure"="OPRENAFL",
         
         "Total operation time"="OPTIME", "Sodium"="PRSODM1",
         "BUN"="PRBUN1",
         "Serum Creatinine"="PRCREAT1",
         "Serum Albumin"="PRALBUM1",
         "Total Bilirubin"="PRBILI1",
         "SGOT"="PRSGOT1",
         "Alkaline Phosphatase"="PRALKPH1",
         "WBC"="PRWBC1",
         "Hematocrit"="PRHCT1",
         "Platelet count"="PRPLATE1",
         "PTT"="PRPTT1",
         "INR"="PRINR1"  )


```



# Data visualization

*1. Make the demographic table*
```{r}
demographics <- dt_renamed %>%
  select("Sex","Race",
         "Ethnicity",  "Smoke",
         "Diabetes", "COPD",
         "Dyspnea", "Congestive Heart Failure",
         "Hypertension", "Ascites", 
         "Disseminated Cancer","Bleeding disorders",
         "Severe Acure Renal Failure")%>%
  tbl_summary(by=  "Severe Acure Renal Failure",
              sort = list(everything() ~ "frequency"))%>%
  add_overall()

demographics

```

*2. Make the summary of the lab results*

```{r}
lab <- dt_renamed %>%
  select( "Age", "BMI","Total operation time","Sodium",
          "BUN",
          "Serum Creatinine",
          "Serum Albumin",
          "Total Bilirubin",
          "SGOT",
          "Alkaline Phosphatase",
          "WBC",
          "Hematocrit",
          "Platelet count",
          "PTT",
          "INR", "Severe Acure Renal Failure")%>%
  tbl_summary(by= "Severe Acure Renal Failure",
              missing = "no",
              digits = list(everything() ~ c(1, 1)),
              statistic = list(all_continuous() ~ "{mean} ({sd}) / ({max}-{min})"))%>%
  add_overall()

lab
```
